\documentclass[
  10pt,
  a4paper,
  twoside,
  openright,
]{book}

\title{Theory and Proof-of-Concept Implementations \\ of Streaming Algorithms for Multiset Data}
\author{Martin de Spirlet \\ 1785605}
\date{September 2021}

\newcommand*{\degree}{MSc Computer Science}
\newcommand*{\supervisor}{Dr.\@ Rajesh Chitnis}
\newcommand*{\school}{School of Computer Science \\ University of Birmingham}

% Set document geometry
\usepackage[a4paper]{geometry} % Flexible and complete interface to document dimensions

\usepackage[chapter]{algorithm} % Floating `algorithm' environment
\usepackage{algpseudocode} % Pseudocode environments for use with the `algorithmicx' style
\usepackage{amsfonts} % TeX fonts from the American Mathematical Society
% \usepackage{amsmath} % AMS mathematical facilities for LaTeX
\usepackage{amssymb} % TeX fonts from the American Mathematical Society
\usepackage{booktabs} % Publication quality tables in LaTeX
\usepackage{calc} % Simple arithmetic in LaTeX commands
\usepackage{centernot} % Centred \not command
\usepackage[inline]{enumitem} % Control layout of itemize, enumerate, description
\usepackage{fancyhdr} % Extensive control of page headers and footers in LaTeX2e
\usepackage{float} % Improved interface for floating objects
\usepackage[edges]{forest} % Drawing (linguistic) trees
\usepackage[htt]{hyphenat} % Disable/enable hypenation
\usepackage{listings} % Typeset source code listings using LaTeX
\usepackage[tbtags]{mathtools} % Mathematical tools to use with amsmath
\usepackage[numbers]{natbib} % Flexible bibliography support
\usepackage[
  % skip = 1\baselineskip plus 2pt,
  skip = 1\baselineskip minus 2pt,
  indent,
]{parskip} % Layout with zero \parindent, non-zero \parskip
\usepackage[simplified]{pgf-umlcd} % Some LaTeX macros for UML Class Diagrams
\usepackage{pgfplots} % Create normal/logarithmic plots in two and three dimensions
\usepackage[onehalfspacing]{setspace} % Set space between lines % setspace must be loaded before hyperref
\usepackage{siunitx} % A comprehensive (SI) units package
\usepackage{stmaryrd} % St Mary Road symbols for theoretical computer science

\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref} % Extensive support for hypertext in LaTeX

\usepackage[
  capitalise,
  nameinlink,
]{cleveref} % Intelligent cross-referencing % cleveref must be loaded after hyperref
\usepackage{usebib} % A simple bibliography processor % usebib must be loaded after hyperref

% Set the reference names for equations (using cleveref)
\Crefname{equation}{Equation}{Equations}
\crefname{equation}{eq.}{eqs.}

% Set the list style (using enumitem)
\setlist[enumerate]{
  label = {\arabic*)},
}

% Set the fancy page style (using fancyhdr)
\fancyhf{}
\fancyhead[LE]{\leftmark}
\fancyhead[RO]{\rightmark}
\fancyfoot[C]{\thepage}
\renewcommand*{\headrulewidth}{0pt}

% Set the default placement specifier for algorithms (using float)
\floatplacement{algorithm}{tbp}

% Set the listing style (using listings)
\lstset{
  basicstyle = \ttfamily,
}
\renewcommand*{\lstinline}[1]{\texttt{#1}}

% Set the bibliography style (using natbib)
\bibliographystyle{IEEEtranN}

% Set the UML diagram colours (using pgf-umlcd)
\renewcommand*{\umldrawcolor}{black}
\renewcommand*{\umlfillcolor}{none}
\renewcommand*{\umltextcolor}{black}

% Use bibliography fields in the document (using usebib)
\bibinput{bibliography}

% Increase penalty for widows and orphans (maximum 10000)
\widowpenalty = 10000
\clubpenalty = 10000

% Include command and environment definitions
\input{definitions/algorithm.tex}
\input{definitions/math.tex}
\input{definitions/structure.tex}

\begin{document}

\maketitle

\pagestyle{fancy}

\frontmatter

\begin{frontsection}{Abstract}
  The number of global sources of data and the volume of information they produce are growing at an exponential rate.
  In fact, the global capacity for the generation of data is increasing at a far greater rate than the global capacity for its storage.
  The storage and processing of such increasingly vast volumes of `big data' presents obvious problems, due to limits on what is economically feasible to store for any extended period of time.
  One solution to the problem of processing big data is the paradigm of \emph{streaming algorithms}.
  A streaming algorithm accepts data one datum at a time from a stream of arbitrary length and arbitrary order, and constructs a compact summary of that data.
  By considering only the current state of the summary and the next datum from the stream during an update, a summary of constantly changing data can be maintained in real time.
  In general, streaming algorithms offer a large saving of space in exchange for a small loss of accuracy.
  In this project, three summaries of multiset data are investigated.
  These are the fingerprint summary, the count sketch and the dyadic count sketch.
  Of these, the fingerprint summary and dyadic count sketch have not attracted many implementations conveniently available online.
  Each corresponding set of algorithms is understood theoretically, expressed formally in pseudocode, and analysed in order to reason about its correctness.
  The algorithms are also implemented in Java to create an extensible library as proof of concept.
  The implementations are then evaluated on artificial datasets carefully designed to test the accuracy and performance of the summaries.
  The results provide empirical verification of the correctness of the summaries and the streaming algorithms that produce them.
  Overall, the project shows how streaming algorithms can be implemented, and provides evidence of their accuracy and performance.
\end{frontsection}

\tableofcontents

\mainmatter

\input{chapters/introduction.tex}

\input{chapters/fingerprint.tex}

\input{chapters/count-sketch.tex}

\addtocontents{toc}{\protect\newpage}

\input{chapters/dyadic-count-sketch.tex}

\input{chapters/discussion.tex}

\input{chapters/conclusions.tex}

\appendix

\input{appendices/repository.tex}

\input{appendices/notation.tex}

\backmatter

\refstepcounter{chapter}
\addcontentsline{toc}{chapter}{\bibname}
\bibliography{bibliography}

\end{document}
